---
title: "Stage 2: Experiment Design Summary"
author: "Bo Yu, Si Liu, Wei Sun"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = 1)
```

## Load experiment object

```{r}
library(jaxExperiment)

expt_file = file.path("outputs", "experiment.rds")
expt = load_experiment(expt_file)
expt
```

## Build experiment design summaries

```{r}
spec2 = stage2_spec_default()
rep2 = stage2_report(expt, spec = spec2)
```

## Persist design fields on experiment object

```{r}
expt = stage2_build_design(expt, spec = spec2)
save_experiment(expt, expt_file)
```

### Columns used

```{r}
cols_used = rep2$columns_used
cols_used_df = data.frame(
  name = names(cols_used),
  value = vapply(cols_used, \(x) if (is.null(x)) "" else as.character(x), character(1)),
  stringsAsFactors = FALSE
)
knitr::kable(cols_used_df, row.names = FALSE)
```

## Marginal summaries

### Model systems

```{r}
knitr::kable(rep2$model_system_summary, row.names = FALSE)
```

### Oxygen condition

```{r}
knitr::kable(rep2$condition_summary, row.names = FALSE)
```

### Time points (day)

```{r}
knitr::kable(rep2$timepoint_summary, row.names = FALSE)
```

### Runs

```{r}
knitr::kable(rep2$run_summary, row.names = FALSE)
```

### Perturbation strategies

```{r}
knitr::kable(rep2$ko_strategy_summary, row.names = FALSE)
```

### Perturbed genes (KO only)

```{r}
knitr::kable(rep2$ko_gene_summary, row.names = FALSE)
```

## Gene-group (WT control mapping)

### Gene-group map (from WT control descriptions)

```{r}
knitr::kable(rep2$gene_group_map, row.names = FALSE)
```

### Sample counts by gene-group

```{r}
knitr::kable(rep2$gene_group_summary, row.names = FALSE)
```

### WT clone controls

```{r}
knitr::kable(rep2$wt_control_map, row.names = FALSE)
```

## DE groups and composition

```{r}
knitr::kable(rep2$de_group_summary, row.names = FALSE)
```

### Within each DE group: WT and KO sample sizes

```{r}
knitr::kable(rep2$de_group_gene_summary, row.names = FALSE)
```

```{r}
gc()
sessionInfo()
```
