---
title: "Differential expression analysis for 2024-06 release JAX_RNAseq4_ExEm_brain"
author: "Si Liu, Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Differential expression using the JAX_RNAseq4_ExEm_brain bulk RNAseq data from the June release.

## R libraries.

```{r load_libraries, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(stringr)
library(MASS)
library(RColorBrewer)
library(DESeq2)

library(viridis)
library(ggpointdensity)
library(dplyr)
library(data.table)
library(ComplexHeatmap)
library(UpSetR)
library(readxl)
theme_set(theme_classic())

path = "../../MorPhiC/data/June-2024/JAX_RNAseq4_ExEm_brain/"
dat_path = paste0(path, "processed/release110-gencode44/Tables")
```


## Read in meta data
```{r fig.width = 3, fig.height=3}
meta = fread("data/June_2024/JAX_RNAseq4_ExEm_brain_meta_data.tsv", 
             data.table = FALSE)
dim(meta)
meta[1:2,]
names(meta)

table(table(meta$biomaterial_id))
table(table(meta$lib_biomaterial_id))
table(table(meta$differentiated_biomaterial_id))
```


## Read in gene count data and filter genes

```{r fig.width = 4, fig.height=3}
cts = fread(file.path(dat_path, "genesCounts.csv"), data.table = FALSE)
dim(cts)
cts[1:2, c(1:2, (ncol(cts)-1):ncol(cts))]

stopifnot(setequal(names(cts)[-1], meta$file_id))

meta = meta[match(names(cts)[-1], meta$file_id),]
table(names(cts)[-1] == meta$file_id)

cts_dat = data.matrix(cts[,-1])
rownames(cts_dat) = cts$Name

table(meta$model_organ, meta$differentiated_timepoint_value)

unique_timepoints = unique(meta$differentiated_timepoint_value)
unique_timepoints = sort(unique_timepoints)

meta$differentiated_timepoint_value = as.character(meta$differentiated_timepoint_value)
meta$differentiated_timepoint_value = factor(meta$differentiated_timepoint_value, 
                                             levels = as.character(unique_timepoints))
```

### Discard genes whose expression is 0 in more than 75% of samples

```{r fig.width = 4, fig.height=3}

model_s = meta$model_system
table(model_s, useNA="ifany")

get_q75 <- function(x){quantile(x, probs = 0.75)}

# there are two levels for model_system
# the result from apply(cts_dat, 1, tapply, model_s, min) is a matrix
# do transpose to the results from apply(cts_dat, 1, tapply, model_s, min)
gene_info = data.frame(Name = cts$Name, 
                       t(apply(cts_dat, 1, tapply, model_s, min)), 
                       t(apply(cts_dat, 1, tapply, model_s, median)), 
                       t(apply(cts_dat, 1, tapply, model_s, get_q75)), 
                       t(apply(cts_dat, 1, tapply, model_s, max)))

dim(gene_info)
gene_info[1:2, ]
table(row.names(gene_info)==gene_info$Name)

names(gene_info)[2:9] = paste0(rep(c("CBO_", "PrS_"), times=4), 
                               rep(c("min", "median", "q75", "max"), each=2))
dim(gene_info)
gene_info[1:2,]

summary(gene_info[,-1])

table(gene_info$CBO_q75 > 0, gene_info$PrS_q75 > 0)

w2kp = which(gene_info$CBO_q75 > 0 | gene_info$PrS_q75 > 0)
cts_dat = cts_dat[w2kp,]
dim(cts_dat)

gene_info = gene_info[w2kp,]

meta$read_depth = colSums(cts_dat)/1e6
meta$q75 = apply(cts_dat, 2, quantile, probs = 0.75)

# there are more than one model systems here
# need to set shape using model_system
ggplot(meta, aes(x=read_depth, y=q75, color=ko_strategy, shape = model_system)) +
    geom_point(size=2, alpha=0.7) + ggtitle("Gene counts") + 
  scale_shape_manual(values = c(7, 16)) + 
  xlab("read depth (million)") + ylab("75 percentile of gene expression") + 
  labs(color = "KO type", shape = "Model") + 
  scale_color_brewer(palette = "Set1")

meta$run_id_short = str_replace(meta$run_id, "^[^-]*-", "")
table(meta$model_system, meta$run_id_short)
ggplot(meta, aes(x=read_depth, y=q75, color=run_id_short, shape = model_system)) +
    geom_point(size=2, alpha=0.7) + ggtitle("Gene counts") + 
  scale_shape_manual(values = c(7, 16)) + 
  xlab("read depth (million)") + ylab("75 percentile of gene expression") + 
  labs(color = "Run ID", shape = "Model") + 
  scale_color_brewer(palette = "Set1")

```

# Check the effect of run id among WT

```{r fig.width = 6, fig.height=4.5}
sample2kp = which(meta$ko_gene == "WT")
cts_dat_m = cts_dat[,sample2kp]
meta_m    = meta[sample2kp,]
summary(meta_m$q75)
dim(cts_dat_m)

# as of R Version 4.1.2 on Hutch server, names(cts_dat_m) does not give the column names of cts_dat_m
# use colnames(cts_dat_m) instead
stopifnot(all(meta_m$file_id == colnames(cts_dat_m)))
table(meta_m$file_id == colnames(cts_dat_m), useNA="ifany")

q75 = apply(cts_dat_m, 1, quantile, probs=0.75)
cts_dat_n = t(cts_dat_m[q75 > 0,])
cts_dat_n = log(median(meta_m$q75)*cts_dat_n/meta_m$q75 + 1)
dim(cts_dat_n)

sample_pca = prcomp(cts_dat_n, center = TRUE, scale. = TRUE)
names(sample_pca)
pc_scores = sample_pca$x
eigen_vals = sample_pca$sdev^2
eigen_vals[1:5]/sum(eigen_vals)
pvs = round(eigen_vals[1:5]/sum(eigen_vals)*100,1)
pvs

PC_df = data.frame(pc_scores)
PC_df = cbind(PC_df, meta_m)

# there are two levels for model_system here
# need to use model_system for the shape
gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=model_system,
                        color=run_id_short)) +
  geom_point(size=2.5, alpha=0.7) + 
  #scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(7, 16)) + 
  labs(color="KO gene", shape ="KO strategy") + 
  xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
  ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
  ggtitle("Wild type samples") + 
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  ) + 
  theme_classic() + 
  theme(
    legend.margin = margin(0, 0, 0, 0), 
    legend.box.just = "left", 
    legend.direction = "vertical" 
  )

print(gPC)

table(PC_df$run_id, PC_df$differentiated_timepoint_value)

gPC_time = ggplot(PC_df, aes(x = PC1, y = PC2, shape=model_system,
                            color=differentiated_timepoint_value)) +
  geom_point(size=2.5, alpha=0.7) + 
  #scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(7, 16)) + 
  labs(color="KO gene", shape ="KO strategy") + 
  xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
  ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
  ggtitle("Wild type samples") + 
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  ) + 
  theme_classic() + 
  theme(
    legend.margin = margin(0, 0, 0, 0), 
    legend.box.just = "left", 
    legend.direction = "vertical" 
  )

print(gPC_time)


```

# Analyze samples of different model systems

For the June 2024 release of JAX_RNAseq4_ExEm_brain, there are two model systems. 

There is only one gene knock out strategy.

```{r fig.width = 6, fig.height=4.5, message=FALSE, warning=FALSE}
table(meta$run_id, meta$model_system)
table(meta$run_id, meta$ko_gene)
table(meta$ko_strategy, meta$ko_gene)

unique_model_ss = unique(meta$model_system)
unique_model_ss = sort(unique_model_ss)

for(model1 in unique_model_ss){
  print("==========================================")
  print(model1)
  print("==========================================")
  sample2kp = which(meta$model_system == model1)
  cts_dat_m = cts_dat[,sample2kp]
  meta_m    = meta[sample2kp,]
 
  # the line right below cannot be used since colnames(cts_dat_m) 
  # for this data file does not have clean pattern
  #table(meta_m$ko_strategy, substr(colnames(cts_dat_m), 1, 7))
  table(meta_m$ko_strategy)
  stopifnot(all(meta_m$file_id == colnames(cts_dat_m)))
  table(meta_m$file_id == colnames(cts_dat_m), useNA="ifany")
  q75 = apply(cts_dat_m, 1, quantile, probs=0.75)

  cts_dat_n = t(cts_dat_m[q75 > 0,])
  cts_dat_n = log(median(meta_m$q75)*cts_dat_n/meta_m$q75 + 1)
  dim(cts_dat_n)
  
  sample_pca = prcomp(cts_dat_n, center = TRUE, scale. = TRUE)
  names(sample_pca)
  pc_scores = sample_pca$x
  eigen_vals = sample_pca$sdev^2
  eigen_vals[1:5]/sum(eigen_vals)
  pvs = round(eigen_vals[1:5]/sum(eigen_vals)*100,1)
  pvs
  
  PC_df = data.frame(pc_scores)
  PC_df = cbind(PC_df, meta_m)
  
  colnames(PC_df)[which(colnames(PC_df)=="differentiated_timepoint_value")] = "day"
  
  only_ko_strategy = setdiff(unique(PC_df$ko_strategy), "WT")
  
  print(length(unique(PC_df$ko_gene)))
  print(length(unique(PC_df$run_id_short)))
  print(length(unique(PC_df$day)))
  print(table(meta_m$run_id, meta_m$ko_gene))

  unique_ko_genes = sort(unique(PC_df$ko_gene))
  unique_ko_genes = c(unique_ko_genes[unique_ko_genes!="WT"], "WT")
  PC_df$ko_gene = factor(PC_df$ko_gene, unique_ko_genes)
  
  gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=ko_gene, color=day)) +
    geom_point(size=2.5, alpha=0.7) + scale_color_brewer(palette = "Dark2") +
      scale_shape_manual(values = c(1, 7, 8, 9, 10, 11, 15, 16, 17)[1:length(unique(PC_df$ko_gene))]) + 
    xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
    ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
    ggtitle(paste0(model1, " ", only_ko_strategy, " for knockout genes")) + 
    guides(
      color = guide_legend(title = NULL, order = 1),
      shape = guide_legend(title = NULL, order = 2)
    ) + 
    theme_classic() + 
    theme(
      legend.margin = margin(0, 0, 0, 0), 
      legend.box.just = "left", 
      legend.direction = "vertical" 
    )

  print(gPC)
  
  gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=run_id_short, color=day)) +
    geom_point(size=2.5, alpha=0.7) + 
    scale_color_brewer(palette = "Dark2") +
    scale_shape_manual(values = c(1, 7, 8, 9, 10, 11, 15, 16, 17)[1:length(unique(PC_df$run_id_short))]) + 
    xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
    ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
    ggtitle(paste0(model1, " ", only_ko_strategy, " for knockout genes")) + 
    guides(
      color = guide_legend(title = NULL, order = 1),
      shape = guide_legend(title = NULL, order = 2)
    ) +
    theme_classic() + 
    theme(
      legend.margin = margin(0, 0, 0, 0), 
      legend.box.just = "left", 
      legend.direction = "vertical" 
    )

  print(gPC)
  
  # day 32 only has 1 WT, not going to run DESeq2 on it
  print(table(meta_m$run_id, meta_m$ko_gene))
  
  gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=run_id, color=ko_gene)) +
  geom_point(size=2.5, alpha=0.7) + 
  scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(1, 7, 8, 9, 10, 11, 15, 16, 17)[1:length(unique(PC_df$run_id_short))]) + 
  xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
  ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
  ggtitle(paste0(model1, " ", only_ko_strategy, " for knockout genes")) + 
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  ) +
  theme_classic() + 
  theme(
    legend.margin = margin(0, 0, 0, 0), 
    legend.box.just = "left", 
    legend.direction = "vertical" 
  )

  print(gPC)

  
  ## Generate sample information matrix
  
  cols2kp = c("ko_strategy", "ko_gene", "run_id", "run_id_short", "differentiated_timepoint_value", "q75")
  sample_info = meta_m[,cols2kp]
  colnames(sample_info)[which(colnames(sample_info)=="differentiated_timepoint_value")] = "day"
  
  dim(sample_info)
  sample_info[1:2,]
  
  table(sample_info$ko_strategy, sample_info$ko_gene)
  
  sample_info$ko_group = paste(sample_info$ko_gene, 
                               sample_info$ko_strategy, sep="_")
  print(table(sample_info$ko_group))
  print(length(unique(sample_info$ko_group)))
  print(table(sample_info$run_id, sample_info$ko_group))
  
  print(table(sample_info$ko_group, sample_info$day))
  
  sample_info$ko_group_run_id = paste(sample_info$ko_group, 
                                      sample_info$run_id_short, sep="_")

  print(table(sample_info$ko_gene, sample_info$day)) 
  print(table(sample_info$ko_group_run_id, sample_info$day)) 
  
  
  # print the table of ko_group and run_id by day
  unique_timepoints_model_s = unique(sample_info$day)
  unique_timepoints_model_s = as.character(sort(unique_timepoints_model_s))
  
  for (cur_day in as.character(unique_timepoints_model_s)){
    sample_info_by_day = sample_info[which(sample_info$day==as.character(cur_day)), ]
    print("---------------------------")
    print(paste0("Model system: ", model1))
    print(paste0("On day ", as.character(cur_day), ": "))
    print(table(sample_info_by_day$ko_group, sample_info_by_day$run_id_short))     
  }
  
  sample_info$log_q75 = log(sample_info$q75)
  
  
  # run DESeq2 for different day groups separately
  # For model system CBO:
  # day 9
  # day 32 (cannot do it, since there is only one WT sample)
  # For model system PrS:
  # day 6
  # do not include run_id in the model if there is only one run_id in certain day

  for (d_group in unique_timepoints_model_s){

    dg_index = which(sample_info$day==d_group)
    cts_dat_m_dg = cts_dat_m[, dg_index]
    sample_info_dg = sample_info[dg_index, ]
    
    stopifnot(all(sample_info_dg$day==d_group))
    
    print("-----------------------------------")
    print(paste0("Model system: ", model1))  
    if (sum(sample_info_dg$ko_gene=="WT")<2){
      print(sprintf("day group %s has only %d wild type samples", 
                    d_group, sum(sample_info_dg$ko_gene=="WT")))
      print(sprintf("do not run DESeq2 for it"))
      print("-----------------------------------")  
      break
    }else{
      print(sprintf("day group %s DESeq2 results:", d_group))
      print("-----------------------------------")  
    }    
    
    
    if (length(unique(sample_info_dg$run_id))==1){
      dds = DESeqDataSetFromMatrix(cts_dat_m_dg, colData=sample_info_dg, 
                                    ~ ko_group + log_q75)
    }else{
      dds = DESeqDataSetFromMatrix(cts_dat_m_dg, colData=sample_info_dg, 
                                    ~ ko_group + run_id + log_q75)     
    }
  
    start.time <- Sys.time()
    dds = DESeq(dds)
    end.time <- Sys.time()
      
    time.taken <- end.time - start.time
    print(time.taken)
    
    ## association with read-depth
    res_rd = results(dds, name = "log_q75")
    res_rd = as.data.frame(res_rd)
    dim(res_rd)
    res_rd[1:2,]
    
    table(is.na(res_rd$padj))
  
    g0 = ggplot(res_rd %>% subset(!is.na(padj)), aes(x=pvalue)) + 
      geom_histogram(color="darkblue", fill="lightblue", 
                     breaks = seq(0,1,by=0.01)) + 
      ggtitle(paste0(model1, " day group ", d_group, " read depth"))
    print(g0)
  
    ## association with run_id
      
    if (length(unique(sample_info_dg$run_id))>1){
      
      dds_lrt = DESeq(dds, test="LRT", reduced = ~ ko_group + log_q75)
      res_lrt = results(dds_lrt)
    
      res_run_id = as.data.frame(res_lrt)
      dim(res_run_id)
      res_run_id[1:2,]
    
      table(is.na(res_run_id$padj))
      
      g0 = ggplot(res_run_id %>% subset(!is.na(padj)), aes(x=pvalue)) + 
        geom_histogram(color="darkblue", fill="lightblue", 
                       breaks = seq(0,1,by=0.01)) + 
        ggtitle(paste0(model1, " day group ", d_group, " Run ID"))
      print(g0)
    
    }
    
    ## DE analysis for each KO gene
    table(sample_info_dg$ko_gene)
    
    genos = setdiff(unique(sample_info_dg$ko_gene), "WT")
    genos
    
    ko_grp  = c("PTC")
    res_geno_df = NULL
    
    for(geno in genos){
      
      res_geno = list()
      
      for(k_grp1 in ko_grp){
        
        res_g1 = results(dds, contrast = c("ko_group", 
                                           paste0(geno, "_", k_grp1), "WT_WT"))
        res_g1 = signif(data.frame(res_g1), 3)
        res_geno[[k_grp1]] = res_g1
        
        g1 = ggplot(res_g1 %>% subset(!is.na(padj)), aes(x=pvalue)) + 
          geom_histogram(color="darkblue", fill="lightblue", 
                         breaks = seq(0,1,by=0.01)) + 
          ggtitle(paste0(model1, " day group ", d_group, " ", geno, "_", k_grp1))
        print(g1)
  
        tag1 = sprintf("%s_%s_%s_%s_", model1, d_group, geno, k_grp1)
        colnames(res_g1) = paste0(tag1, colnames(res_g1))
        
        if(is.null(res_geno_df)){
          res_geno_df = res_g1
        }else if(!is.null(res_geno_df)){
          stopifnot(all(rownames(res_geno_df) == rownames(res_g1)))
          res_geno_df = cbind(res_geno_df, res_g1)
        }
      }
      
      get_index <- function(x){
        which(x$padj < 0.05 & abs(x$log2FoldChange) > log2(1.5))
      }
      
      w_ptc = get_index(res_geno[["PTC"]])
  
      print(paste0(model1, " day group ", d_group, " under KO gene ", geno, ":"))
      
      print(paste0(model1, " number of DE genes from knock out strategy PTC: ", 
                   as.character(length(w_ptc))))
            
    }
  
    dim(res_geno_df)
    res_geno_df[1:2,]
    
    res_df = data.frame(gene_ID=rownames(res_geno_df), res_geno_df)
    dim(res_df)
    res_df[1:2,]

    output_dir = "results/2024_06_JAX_RNAseq4_ExEm_brain"
    
    if (!file.exists(output_dir)){
      dir.create(output_dir)
    }
  
    fwrite(res_df, 
           sprintf("%s/2024_06_JAX_RNAseq4_ExEm_brain_%s_%s_DEseq2.tsv", 
                   output_dir, model1, d_group), 
           sep="\t")
    
  }
  
}


```

```{r}
gc()
sessionInfo()
```

