---
title: "Differential expression analysis for 2026-01 release MSK_ISL1_OE"
author: "Si Liu, Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Differential expression using the bulk RNAseq MSK_ISL1_OE the Jan. 2026 release.

This dataset spans

* H1,
* day18, (final_timepoint: "unknown"),
* four different colonal cell lines,
* pancreatic islet,
* OE of BFP only as control, and OE of ISL1 as case. 

The test of DE analysis planned:

* for each colonal cell line, ISL1 OE vs control
* for baseline, the other three colonal cell lines vs the WT cell line

## R libraries.

```{r load_libraries, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(stringr)
library(MASS)
library(RColorBrewer)
library(DESeq2)

library(viridis)
library(ggpointdensity)
library(dplyr)
library(data.table)
library(ComplexHeatmap)
library(UpSetR)
library(readxl)
theme_set(theme_classic())

path = "../../MorPhiC/data/MorPhiC_internal_releases/MorPhiC_Release_January_2026/MSK_bulkRNAseq_ISL1_OE"
dat_path = file.path(path, "processed/Tables")

meta_save_path = "./data/Jan_2026"
meta_file = file.path(meta_save_path, "MSK_bulkRNAseq_ISL1_OE_meta_data.tsv")
  
qc_path = "./qc_figures"
# create the figure path if it does not exist
dir.create(qc_path, showWarnings = FALSE, recursive = TRUE)

```

## Load processed meta data table
```{r fig.width = 3, fig.height=3}
meta = fread(meta_file, sep="\t", data.table = FALSE)
dim(meta)
```


## Read in gene count data and filter genes

There are redundancy in the count data file.

    The 2nd and 3rd columns are the same and only one of them should be kept. 
    Only keep the columns that show up in the sequence table

```{r fig.width = 4, fig.height=3}
cts_lanes = fread(file.path(dat_path, "genesCounts.csv"), data.table = FALSE)
dim(cts_lanes)
cts_lanes[1:2, c(1:2, (ncol(cts_lanes)-1):ncol(cts_lanes))]

names(cts_lanes)[1:6]

length(unique(colnames(cts_lanes)[2:ncol(cts_lanes)]))

# remove the third column from the cts_lanes object since it is a duplicate of the second column
cts_lanes = cts_lanes[,-3]
names(cts_lanes)[1:6]

# only keep the columns that show up in the sequence table
# summarized as the info in the sample_id column of meta data table
names_short = gsub("_L00[1-8]$", "", names(cts_lanes))
cols2kp = which(names_short %in% meta$sample_id)
length(cols2kp)
cts_lanes = cts_lanes[, c(1, cols2kp)]
dim(cts_lanes)

names_short_filterd = gsub("_L00[1-8]$", "", names(cts_lanes))

stopifnot(setequal(names_short_filterd[-1], meta$sample_id))

# Separate the gene ID column from the count columns
gene_column <- cts_lanes$Name
counts   <- as.matrix(cts_lanes[, -1])   # drop the "Name" column

# Derive sample ID from each column name by stripping "_L00X" at the end
sample_id <- sub("_L00[1-8]$", "", colnames(counts))

# Sum lane columns belonging to the same sample.
collapsed <- t(rowsum(t(counts), group = sample_id))

# shorten column names further to just "ISL1-BFP-1" etc.
colnames(collapsed) <- sub("_IGO.*", "", colnames(collapsed))

# Put the gene IDs back as a column
cts <- data.frame(Name = gene_column, collapsed, check.names = FALSE)

head(cts)
dim(cts_lanes)        # original
dim(cts)   # should have ~1/4 the columns + 1 for Name

# reorder the meta data table to match the samples in the processed count data
meta = meta[match(names(cts)[-1], meta$clonal.label),]
table(names(cts)[-1] == meta$clonal.label, useNA="ifany")

cts_dat = data.matrix(cts[,-1])
rownames(cts_dat) = cts$Name

table(meta$timepoint_value, useNA="ifany")
meta$timepoint = as.character(meta$timepoint_value)

```

### Read in gene annoation
```{r}
gene_anno = fread("data/gencode_v44_primary_assembly_info.tsv")
dim(gene_anno)
gene_anno[1:2,]

table(rownames(cts_dat) %in% gene_anno$ensembl_gene_id)
# all ensembl gene IDs are contained in the annotation
setdiff(rownames(cts_dat), gene_anno$ensembl_gene_id)
```

### Discard genes whose expression is 0 in more than 75% of samples

```{r fig.width = 4, fig.height=3}
model_s = meta$model_system
table(model_s, useNA="ifany")

get_q75 <- function(x){quantile(x, probs = 0.75)}

gene_info = data.frame(Name = cts$Name,
                       apply(cts_dat, 1, min), 
                       apply(cts_dat, 1, median), 
                       apply(cts_dat, 1, get_q75), 
                       apply(cts_dat, 1, max))

dim(gene_info)
gene_info[1:2,]
table(row.names(gene_info)==gene_info$Name, useNA="ifany")

names(gene_info)[2:5] = paste0(rep(c(model_s), times=4), 
                               rep(c("min", "median", "q75", "max"), each=1))
dim(gene_info)
gene_info[1:2,]

summary(gene_info[,-1])

w2kp = which(gene_info$Pancreatic_Isletq75 > 0)
cts_dat = cts_dat[w2kp,]
dim(cts_dat)

gene_info = gene_info[w2kp,]

meta$read_depth = colSums(cts_dat)/1e6
meta$q75 = apply(cts_dat, 2, quantile, probs = 0.75)
meta$detected_genes = colSums(cts_dat > 0)
meta$top_gene_fraction = apply(cts_dat, 2, function(x) max(x)/sum(x))
meta$top20_fraction = apply(cts_dat, 2, function(x) sum(sort(x, decreasing = TRUE)[1:20])/sum(x))
meta$q75_outlier = meta$read_depth > 50 & meta$q75 < 1500

# add label to the two points with read_depth above 50 and q75 below 1500

g_counts = ggplot(meta, aes(x = read_depth, y = q75, color = cell_line, shape = OE_status)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text_repel(
    data = subset(meta, read_depth > 50 & q75 < 1500),
    aes(label = clonal.label),     # whichever column holds the sample name
    size = 3,
    show.legend = FALSE,           # don't add a label glyph to the legend
    max.overlaps = Inf,            # don't silently drop labels if crowded
    box.padding = 0.4,
    point.padding = 0.3,
    min.segment.length = 0
  ) +
  ggtitle("Gene counts") +
  xlab("read depth (million)") +
  ylab("75 percentile of gene expression") +
  scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(7, 15, 16, 17)) +
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  )

print(g_counts)

g_detected_genes = ggplot(meta, aes(x = detected_genes, y = q75,
                                    color = cell_line, shape = OE_status)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text_repel(
    data = subset(meta, q75_outlier),
    aes(label = clonal.label),
    size = 3,
    show.legend = FALSE,
    max.overlaps = Inf,
    box.padding = 0.4,
    point.padding = 0.3,
    min.segment.length = 0
  ) +
  ggtitle("Detected genes") +
  xlab("number of detected genes") +
  ylab("75 percentile of gene expression") +
  scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(7, 15, 16, 17)) +
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  )

print(g_detected_genes)

g_top20_fraction = ggplot(meta, aes(x = top20_fraction, y = q75,
                                    color = cell_line, shape = OE_status)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text_repel(
    data = subset(meta, q75_outlier),
    aes(label = clonal.label),
    size = 3,
    show.legend = FALSE,
    max.overlaps = Inf,
    box.padding = 0.4,
    point.padding = 0.3,
    min.segment.length = 0
  ) +
  ggtitle("Top 20 genes") +
  xlab("fraction of counts in top 20 genes") +
  ylab("75 percentile of gene expression") +
  scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(7, 15, 16, 17)) +
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  )

print(g_top20_fraction)

g_top_gene_fraction = ggplot(meta, aes(x = top_gene_fraction, y = q75,
                                       color = cell_line, shape = OE_status)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_text_repel(
    data = subset(meta, q75_outlier),
    aes(label = clonal.label),
    size = 3,
    show.legend = FALSE,
    max.overlaps = Inf,
    box.padding = 0.4,
    point.padding = 0.3,
    min.segment.length = 0
  ) +
  ggtitle("Top gene") +
  xlab("fraction of counts in top gene") +
  ylab("75 percentile of gene expression") +
  scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(7, 15, 16, 17)) +
  guides(
    color = guide_legend(title = NULL, order = 1),
    shape = guide_legend(title = NULL, order = 2)
  )

print(g_top_gene_fraction)

g_qc_2x2 = ggarrange(g_counts, g_detected_genes, g_top20_fraction, g_top_gene_fraction,
                     ncol = 2, nrow = 2)
qc_pdf = file.path(qc_path, "MSK_ISL1_OE_q75_outlier_qc.pdf")
ggsave(qc_pdf, g_qc_2x2, width = 9, height = 7, units = "in")
#print(g_qc_2x2)


qc_df = meta[, c("clonal.label", "label", "read_depth", "q75", "detected_genes",
                 "top_gene_fraction", "top20_fraction", "q75_outlier")]
head(qc_df[order(qc_df$q75),])

outlier_samples = meta$clonal.label[meta$q75_outlier]
top2_outlier_genes = do.call(rbind, lapply(outlier_samples, function(cur_sample) {
  cur_counts = cts_dat[, cur_sample]
  top2_idx = head(order(cur_counts, decreasing = TRUE), 2)
  data.frame(
    sample = cur_sample,
    gene_id = rownames(cts_dat)[top2_idx],
    hgnc_symbol = gene_anno$hgnc_symbol[match(rownames(cts_dat)[top2_idx],
                                               gene_anno$ensembl_gene_id)],
    count = cur_counts[top2_idx],
    fraction = cur_counts[top2_idx]/sum(cur_counts),
    row.names = NULL
  )
}))

top2_outlier_genes

cts_dat_n_qc = log10(t(cts_dat) + 1)
stopifnot(all(rownames(cts_dat_n_qc) == meta$clonal.label))

plot_top2_gene_pair = function(cur_sample) {
  cur_top2 = top2_outlier_genes[top2_outlier_genes$sample == cur_sample,]
  stopifnot(nrow(cur_top2) == 2)

  cur_df = data.frame(
    gene1_expr = cts_dat_n_qc[, cur_top2$gene_id[1]],
    gene2_expr = cts_dat_n_qc[, cur_top2$gene_id[2]],
    meta
  )

  ggplot(cur_df, aes(x = gene1_expr, y = gene2_expr,
                     color = cell_line, shape = OE_status)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_text_repel(
      data = subset(cur_df, clonal.label %in% outlier_samples),
      aes(label = clonal.label),
      size = 3,
      show.legend = FALSE,
      max.overlaps = Inf,
      box.padding = 0.4,
      point.padding = 0.3,
      min.segment.length = 0
    ) +
    ggtitle(cur_sample) +
    xlab(paste0(cur_top2$hgnc_symbol[1], " log10(count + 1)")) +
    ylab(paste0(cur_top2$hgnc_symbol[2], " log10(count + 1)")) +
    scale_color_brewer(palette = "Dark2") +
    scale_shape_manual(values = c(7, 15, 16, 17)) +
    guides(
      color = guide_legend(title = NULL, order = 1),
      shape = guide_legend(title = NULL, order = 2)
    )
}

g_top2_gene_pairs = lapply(outlier_samples, plot_top2_gene_pair)

cts_dat_n_qc_q75 = t(cts_dat)
cts_dat_n_qc_q75 = log10(median(meta$q75)*cts_dat_n_qc_q75/meta$q75 + 1)

plot_top2_gene_pair_q75_norm = function(cur_sample) {
  cur_top2 = top2_outlier_genes[top2_outlier_genes$sample == cur_sample,]
  stopifnot(nrow(cur_top2) == 2)

  cur_df = data.frame(
    gene1_expr = cts_dat_n_qc_q75[, cur_top2$gene_id[1]],
    gene2_expr = cts_dat_n_qc_q75[, cur_top2$gene_id[2]],
    meta
  )

  ggplot(cur_df, aes(x = gene1_expr, y = gene2_expr,
                     color = cell_line, shape = OE_status)) +
    geom_point(size = 2, alpha = 0.7) +
    geom_text_repel(
      data = subset(cur_df, clonal.label %in% outlier_samples),
      aes(label = clonal.label),
      size = 3,
      show.legend = FALSE,
      max.overlaps = Inf,
      box.padding = 0.4,
      point.padding = 0.3,
      min.segment.length = 0
    ) +
    ggtitle(paste0(cur_sample, ": q75-normalized")) +
    xlab(paste0(cur_top2$hgnc_symbol[1], " cts_dat_n, log10")) +
    ylab(paste0(cur_top2$hgnc_symbol[2], " cts_dat_n, log10")) +
    scale_color_brewer(palette = "Dark2") +
    scale_shape_manual(values = c(7, 15, 16, 17)) +
    guides(
      color = guide_legend(title = NULL, order = 1),
      shape = guide_legend(title = NULL, order = 2)
    )
}

g_top2_gene_pairs_q75_norm = lapply(outlier_samples, plot_top2_gene_pair_q75_norm)

for (i in seq_along(outlier_samples)){
  print(g_top2_gene_pairs[[i]])
  print(g_top2_gene_pairs_q75_norm[[i]])
}

g_top2_gene_pairs_combined = ggarrange(plotlist = c(g_top2_gene_pairs,
                                                    g_top2_gene_pairs_q75_norm),
                                       ncol = 2, nrow = 2)
top2_gene_pair_combined_pdf = file.path(qc_path,
                                        "MSK_ISL1_OE_q75_outlier_top2_gene_pairs_combined.pdf")
ggsave(top2_gene_pair_combined_pdf, g_top2_gene_pairs_combined,
       width = 7, height = 6.6, units = "in")
# print(g_top2_gene_pairs_combined)
  
```

## Analyze samples of different model systems or conditions

Explore the PC plots first, to decide which level to run DESeq2 on. 

Then run the DE analysis. 

Since there is only one day value for each model system, do not include day value in each scatter plot.

```{r fig.width = 4.5, fig.height=3, message=FALSE, warning=FALSE}
meta$model_condition = meta$model_system
table(meta$model_condition)

unique_model_condition_ss = unique(meta$model_condition)
unique_model_condition_ss = sort(unique_model_condition_ss)

scatterplots_to_save = list(g_counts)

for(cur_mc in unique_model_condition_ss){
  
  print(cur_mc)
  sample2kp = which(meta$model_condition == cur_mc)
  cts_dat_m = cts_dat[,sample2kp]
  meta_m    = meta[sample2kp,]

  stopifnot(all(meta_m$clonal.label == colnames(cts_dat_m)))
  q75 = apply(cts_dat_m, 1, quantile, probs=0.75)

  cts_dat_n = t(cts_dat_m[q75 > 0,])
  cts_dat_n = log(median(meta_m$q75)*cts_dat_n/meta_m$q75 + 1)
  dim(cts_dat_n)
  
  sample_pca = prcomp(cts_dat_n, center = TRUE, scale. = TRUE)
  names(sample_pca)
  pc_scores = sample_pca$x
  eigen_vals = sample_pca$sdev^2
  eigen_vals[1:5]/sum(eigen_vals)
  pvs = round(eigen_vals[1:5]/sum(eigen_vals)*100,1)
  pvs

  PC_df = data.frame(pc_scores)
  PC_df = cbind(PC_df, meta_m)
  pc1_rightmost_samples = head(PC_df$clonal.label[order(-PC_df$PC1)], 2)
  
  gPC = ggplot(PC_df, aes(x = PC1, y = PC2, shape=OE_status, color=cell_line)) +
    geom_point(size=2.5, alpha=0.7) + scale_color_brewer(palette = "Dark2") +
    geom_text_repel(
      data = subset(PC_df, PC1 > 100),
      aes(label = clonal.label),     # whichever column holds the sample name
      size = 3,
      show.legend = FALSE,           # don't add a label glyph to the legend
      max.overlaps = Inf,            # don't silently drop labels if crowded
      box.padding = 0.4,
      point.padding = 0.3,
      min.segment.length = 0
    ) + 
      scale_shape_manual(values = c(7, 15, 16, 17)) + 
    xlab(sprintf("PC1 (%.1f%% variance)", pvs[1])) + 
    ylab(sprintf("PC2 (%.1f%% variance)", pvs[2])) + 
    ggtitle(cur_mc) + 
    guides(
      color = guide_legend(title = NULL, order = 1),
      shape = guide_legend(title = NULL, order = 2)
    ) +
    theme(
      legend.margin = margin(0, 0, 0, 0), 
      legend.box.just = "left", 
      legend.direction = "vertical" 
    )

  print(gPC)
  scatterplots_to_save[[length(scatterplots_to_save) + 1]] = gPC

  # All PC1 loadings, named by gene
  pc1_loadings <- sample_pca$rotation[, "PC1"]
  df_pc1_loadings = data.frame(gene_id = names(pc1_loadings), 
                               pc1_loading = pc1_loadings)
  # assign the hgnc_symbol
  df_pc1_loadings$hgnc_symbol = gene_anno$hgnc_symbol[match(df_pc1_loadings$gene_id, 
                                                            gene_anno$ensembl_gene_id)]
  any(duplicated(df_pc1_loadings$gene_id))
  sum(is.na(df_pc1_loadings$hgnc_symbol))
  sum(df_pc1_loadings$hgnc_symbol == "", na.rm = TRUE)
  
  rownames(df_pc1_loadings) = NULL
   
  df_pc1_loadings_top <- df_pc1_loadings[order(-df_pc1_loadings$pc1_loading), ]
  head(df_pc1_loadings_top, 20)
   
  df_pc1_loadings_bottom <- df_pc1_loadings[order(df_pc1_loadings$pc1_loading), ]
  head(df_pc1_loadings_bottom, 20)

  # generate a histogram of PC1 loadings
  g_pc1_loadings = ggplot(df_pc1_loadings, aes(x=pc1_loading)) + 
    geom_histogram(color="darkblue", fill="lightblue", bins = 200) + 
    ggtitle(paste0(cur_mc, " PC1 loadings"))
  g_pc1_loadings
  
  # for each of the genes in df_pc1_loadings_top and df_pc1_loadings_bottom
  # compute the correlation with PC1 values
  cor_top = cor(cts_dat_n[, df_pc1_loadings_top$gene_id[1:20]], pc_scores[, "PC1"])
  cor_bottom = cor(cts_dat_n[, df_pc1_loadings_bottom$gene_id[1:20]], pc_scores[, "PC1"])
  
  as.numeric(cor_top)
  as.numeric(cor_bottom)
  
  # Cell cycle hypothesis — score known cell-cycle genes
  cc_genes <- c("MKI67", "TOP2A", "PCNA", "MCM2", "MCM5", "CCNB1", "CCNA2", "CDK1")
  cc_ids <- gene_anno$ensembl_gene_id[gene_anno$hgnc_symbol %in% cc_genes]
  cc_loadings <- df_pc1_loadings$pc1_loading[df_pc1_loadings$gene_id %in% cc_ids]
  summary(cc_loadings)   # are they systematically positive or negative?
    
  lib_size <- colSums(cts_dat_m)
  cor(sample_pca$x[, "PC1"], lib_size)
  
  oxphos_genes <- c("SDHA","SDHB","SDHC","SDHD",        # Complex II
                    "COX4I1","COX5A","COX5B","COX6A1",  # Complex IV
                    "NDUFA1","NDUFB1","NDUFS1",         # Complex I
                    "ATP5F1A","ATP5F1B","ATP5F1C",      # Complex V
                    "UQCRC1","UQCRC2")                   # Complex III
  oxphos_ids <- gene_anno$ensembl_gene_id[gene_anno$hgnc_symbol %in% oxphos_genes]
  oxphos_loadings <- df_pc1_loadings$pc1_loading[df_pc1_loadings$gene_id %in% oxphos_ids]
  summary(oxphos_loadings)
  
  # PC plot if regressing out the read depth (in the format of q75)
  
  q75_m = apply(cts_dat_m[q75 > 0,], 2, quantile, probs = 0.75)
  stopifnot(all(names(q75_m) == rownames(cts_dat_n)))

  cts_dat_n_log = log(t(cts_dat_m[q75 > 0,]) + 1)
  q75_m_log = log(q75_m + 1)

  cts_dat_q75_resid = apply(cts_dat_n_log, 2, function(expr) {
    residuals(lm(expr ~ q75_m_log))
  })
  rownames(cts_dat_q75_resid) = rownames(cts_dat_n_log)
  resid_gene_sd = apply(cts_dat_q75_resid, 2, sd)
  cts_dat_q75_resid = cts_dat_q75_resid[, resid_gene_sd > 0 & !is.na(resid_gene_sd)]
  dim(cts_dat_q75_resid)

  sample_pca_q75_resid = prcomp(cts_dat_q75_resid, center = TRUE, scale. = TRUE)
  pc_scores_q75_resid = sample_pca_q75_resid$x
  eigen_vals_q75_resid = sample_pca_q75_resid$sdev^2
  pvs_q75_resid = round(eigen_vals_q75_resid[1:5]/sum(eigen_vals_q75_resid)*100, 1)

  PC_df_q75_resid = data.frame(pc_scores_q75_resid)
  PC_df_q75_resid = cbind(PC_df_q75_resid, meta_m)

  gPC_q75_resid = ggplot(PC_df_q75_resid,
                         aes(x = PC1, y = PC2, shape = OE_status, color = cell_line)) +
    geom_point(size = 2.5, alpha = 0.7) +
    scale_color_brewer(palette = "Dark2") +
    geom_text_repel(
      data = subset(PC_df_q75_resid, clonal.label %in% pc1_rightmost_samples),
      aes(label = clonal.label),
      size = 3,
      show.legend = FALSE,
      max.overlaps = Inf,
      box.padding = 0.4,
      point.padding = 0.3,
      min.segment.length = 0
    ) +
    scale_shape_manual(values = c(7, 15, 16, 17)) +
    xlab(sprintf("PC1 (%.1f%% variance)", pvs_q75_resid[1])) +
    ylab(sprintf("PC2 (%.1f%% variance)", pvs_q75_resid[2])) +
    ggtitle(paste0(cur_mc, ": residual after regressing out q75")) +
    guides(
      color = guide_legend(title = NULL, order = 1),
      shape = guide_legend(title = NULL, order = 2)
    ) +
    theme(
      legend.margin = margin(0, 0, 0, 0),
      legend.box.just = "left",
      legend.direction = "vertical"
    )

  print(gPC_q75_resid)
  scatterplots_to_save[[length(scatterplots_to_save) + 1]] = gPC_q75_resid
  
}

g_scatterplots_2x2 = ggarrange(plotlist = scatterplots_to_save, ncol = 2, nrow = 2)
scatterplot_pdf = file.path(qc_path, "MSK_ISL1_OE_exploratory_scatterplots.pdf")
ggsave(scatterplot_pdf, g_scatterplots_2x2, width = 9, height = 7, units = "in")


```


Use each combination of (model system, condition) as DE group. 

Run analysis for each DE group separately.

For the output files, create two versions, one with direct information, one with padj set to NA for the genes when the number of 0 per test is >=4. 

In more details, for both versions of the outputs, make it one file per knockout gene per DE group (DE group can be model system, model system + condition, model system + run_id, model system + day, etc.). 

The first version of the output files contains 
    
    gene id, gene symbol

and for each knockout strategy:

    baseMean, log2FoldChange, lfcSE, stat, pvalue, padj
    
The second version of the output files contains  

    gene id, gene symbol, chr, strand, position

and for each knockout strategy:

    log2FoldChange, pvalue, padj (set to be NA if the number of 0 per test >= 4), 

In addition, save out a separate file of the sample size for each comparison. This needs to be by DE group. 

In this setting, all samples were sequenced in the same batch. Do not consider run.id as a covariate or grouping factor. 

### Prepare output directories for DE results

```{r fig.width = 4.5, fig.height=3, message=FALSE, warning=FALSE}

release_name = "2026_01_MSK_bulkRNAseq_ISL1_OE"

output_dir = file.path("results", release_name)
raw_output_dir = file.path(output_dir, "raw")
processed_output_dir = file.path(output_dir, "processed")

if (!file.exists(raw_output_dir)){
  dir.create(raw_output_dir, recursive = TRUE)
}

if (!file.exists(processed_output_dir)){
  dir.create(processed_output_dir, recursive = TRUE)
}
```

### DE analysis for all genes
```{r fig.width = 4.8, fig.height=3.2, message=FALSE, warning=FALSE}

unique_model_ss = unique(meta$model_system)
unique_model_ss = sort(unique_model_ss)
unique_model_ss

for (model1 in unique_model_ss){
  
  print(model1)
  sample2kp = which(meta$model_system == model1)
  cts_dat_m = cts_dat[,sample2kp]
  meta_m    = meta[sample2kp,]

  stopifnot(all(meta_m$clonal.label == colnames(cts_dat_m)))
  print(table(meta_m$clonal.label == colnames(cts_dat_m), useNA="ifany"))  
  
  ## Generate sample information matrix
  # cols2kp = c("model_system", "condition", "ko_strategy", "ko_gene", "run_id",
  #             "run_id_short", "q75", "timepoint")
  cols2kp = c("model_system", "cell_line", "OE_status", "q75", "timepoint")
  sample_info = meta_m[,cols2kp]
  colnames(sample_info)[which(colnames(sample_info)=="timepoint")] = "day"
    
  dim(sample_info)
  sample_info[1:2,]
      
  print(table(sample_info$cell_line, sample_info$OE_status))
  
  sample_info$ko_group = paste(sample_info$cell_line, 
                               sample_info$OE_status, sep="_")
  print(table(sample_info$ko_group))
  
  # use the combination of (model system, condition) as DE group
  
  sample_info$model_day = paste0(sample_info$model_system, 
                                 "_day_", 
                                 as.character(sample_info$day))
    
  sample_info$de_grp = sample_info$model_day

  sorted_de_grps = sort(unique(sample_info$de_grp))
  sorted_de_grps

  for(d_group in sorted_de_grps){
    
    print(d_group)
    
    dg_index = which(sample_info$de_grp==d_group)
    cts_dat_m_dg = cts_dat_m[, dg_index]
    sample_info_dg = sample_info[dg_index, ]

    stopifnot(all(sample_info_dg$de_grp==d_group))   
    
    print("-----------------------------------")      
    print(paste0("Model system: ", model1))  
    print(sprintf("DE analysis group %s DESeq2 results:", d_group))
    print("-----------------------------------")  
      

    # do two steps of filtering
    # first, filter based on q75 of each gene
    # second, filter based on whether each gene is expressed in at least 4 samples
  
    dg_q75 = apply(cts_dat_m_dg, 1, quantile, probs=0.75)
    cts_dat_m_dg = cts_dat_m_dg[dg_q75 > 0,]
    
    print(sprintf("After filtering by gene expression q75, the number of genes reduces from %d to %d", 
                  length(dg_q75), nrow(cts_dat_m_dg)))
    
    n_pos = rowSums(cts_dat_m_dg>0)
    cts_dat_m_dg = cts_dat_m_dg[n_pos >= 4,]
    
    if (length(n_pos)==nrow(cts_dat_m_dg)){
      print("After filtering by nonzero expression in at least 4 samples, the number of genes does not change.")
    }else{
      print(sprintf("After filtering by nonzero expression in at least 4 samples, the number of genes reduces from %d to %d", 
                    length(n_pos), nrow(cts_dat_m_dg)))    
    }
  
    # update the q75 based on only genes used in the process
    sample_info_dg$q75 = apply(cts_dat_m_dg, 2, quantile, probs = 0.75)
    sample_info_dg$log_q75 = log(sample_info_dg$q75)   
    
    # mean center log_q75 
    sample_info_dg$log_q75 = sample_info_dg$log_q75 - mean(sample_info_dg$log_q75)

    
    # if (length(unique(sample_info_dg$run_id))==1){
      dds = DESeqDataSetFromMatrix(cts_dat_m_dg, colData=sample_info_dg, 
                                    ~ ko_group + log_q75)
    # }else{
    #   dds = DESeqDataSetFromMatrix(cts_dat_m_dg, colData=sample_info_dg, 
    #                                 ~ ko_group + run_id + log_q75)     
    # }
    
    start.time <- Sys.time()
    dds = DESeq(dds)
    end.time <- Sys.time()
    
    time.taken <- end.time - start.time
    print(time.taken)
    
    ## association with read-depth
    res_rd = results(dds, name = "log_q75")
    res_rd = as.data.frame(res_rd)
    dim(res_rd)
    res_rd[1:2,]
    
    table(is.na(res_rd$padj))
    
    g0 = ggplot(res_rd %>% subset(!is.na(padj)), aes(x=pvalue)) + 
      geom_histogram(color="darkblue", fill="lightblue", 
                     breaks = seq(0,1,by=0.01)) + 
      ggtitle(paste0(d_group, " read depth"))
    print(g0)
    
    ## Rerun the analysis without read-depth if it is not significant for 
    ## most genes, and the number of samples is small
    ## i.e., proportion of non-null in the distribution is smaller than 0.01
    ## (this needs to be restricted to the genes with not NA adjusted pvalue)
    ## or if smaller than 0.1 and the number of samples is not greater than 6

    pi_1_rd = max(0, mean(res_rd$pvalue[!is.na(res_rd$padj)] < 0.05) - 0.05)
    pi_1_rd = max(pi_1_rd, 0, 1 - 2*mean(res_rd$pvalue[!is.na(res_rd$padj)] > 0.5))
    
    print(sprintf("prop of non-null for rd: %.2e", pi_1_rd))
    
    if(pi_1_rd < 0.01 || (ncol(cts_dat_m_dg) <= 6 && pi_1_rd < 0.1 )){
      
      print("rerun DESeq2 without read depth")
      
      include_rd = 0
      # if (length(unique(sample_info_dg$run_id))==1){
        dds = DESeqDataSetFromMatrix(cts_dat_m_dg, colData=sample_info_dg, 
                                      ~ ko_group)
      # }else{
      #   dds = DESeqDataSetFromMatrix(cts_dat_m_dg, colData=sample_info_dg, 
      #                                 ~ ko_group + run_id)     
      # }
      start.time <- Sys.time()
      dds = DESeq(dds)
      end.time <- Sys.time()
      
      time.taken <- end.time - start.time
      print(time.taken)
      
    }else{
      include_rd = 1
    }
    
    
     ## association with run_id
    
    # if (length(unique(sample_info_dg$run_id))>1){
    #   
    #   if(include_rd){
    #     dds_lrt = DESeq(dds, test="LRT", reduced = ~ ko_group + log_q75)
    #   }else{
    #     dds_lrt = DESeq(dds, test="LRT", reduced = ~ ko_group)
    #   }
    #   
    #   res_lrt = results(dds_lrt)
    # 
    #   res_run_id = as.data.frame(res_lrt)
    #   dim(res_run_id)
    #   res_run_id[1:2,]
    # 
    #   table(is.na(res_run_id$padj))
    #   
    #   g0 = ggplot(res_run_id %>% subset(!is.na(padj)), aes(x=pvalue)) + 
    #     geom_histogram(color="darkblue", fill="lightblue", 
    #                    breaks = seq(0,1,by=0.01)) + 
    #     ggtitle(paste0(d_group, " Run ID"))
    #   print(g0)
    #   
    # }   
    
    ## DE analysis for two types of comparisons
    ## 1. under each cell line, ISL1 OE vs BFP only
    ## 2. under BFP only, each knockout gene vs WT
    
    table(sample_info_dg$OE_status)
    table(sample_info_dg$OE_status, sample_info_dg$cell_line)
    
    target_OE_genes = "ISL1"
    ko_type = "OE"
    
    cell_lines = unique(sample_info_dg$cell_line)
    
    # 1. under each cell line, ISL1 OE vs BFP only
    
    df_size = NULL
    res_cellline = list()
      
    for(cline in cell_lines){
      
      res_cellline_df = NULL
      res_cellline_reduced_df = NULL
      
      if (cline!="WT"){
        cline_str = paste0(cline, "_KO")
      }else{
        cline_str = "WT"
      }
      
      for(g_oe in target_OE_genes){
        
        res_g1 = results(dds, contrast = c("ko_group", 
                                           paste0(cline, "_", g_oe), paste0(cline, "_BFP")))
        res_g1 = signif(data.frame(res_g1), 3)
        
        # add a column to record the number of zeros per test
        test_index = which(sample_info_dg$ko_group%in%c(paste0(cline, "_", g_oe), paste0(cline, "_BFP")))
        
        cts_dat_m_dg_test = cts_dat_m_dg[, test_index]
        n_zero = rowSums(cts_dat_m_dg_test==0)
        res_g1$n_zeros = n_zero
        
        # record the number of samples involved in the comparison
        test_ko_group_vec = sample_info_dg$ko_group[test_index]
        
        df_size = rbind(df_size, 
                        c(d_group, 
                          cline_str, 
                          paste0(g_oe, "_", ko_type), 
                          sum(test_ko_group_vec!=paste0(cline, "_BFP")), 
                          sum(test_ko_group_vec==paste0(cline, "_BFP"))))
        
        # prepare a processed version of output
        res_g1_reduced = res_g1[, c("log2FoldChange", "pvalue", "padj", "n_zeros")]
        res_g1_reduced$padj[which(res_g1_reduced$n_zeros>=4)] = NA
        
        if ((sum(test_ko_group_vec!=paste0(cline, "_BFP"))==1)|(sum(test_ko_group_vec==paste0(cline, "_BFP"))==1)){
          res_g1_reduced$padj = NA
        }
        
        res_g1_reduced = res_g1_reduced[, c("log2FoldChange", "pvalue", "padj")]
        
        res_cellline[[paste0(cline_str, "_", g_oe, "_", ko_type)]] = res_g1_reduced
        
        if ((sum(test_ko_group_vec!=paste0(cline, "_BFP"))>1) & (sum(test_ko_group_vec==paste0(cline, "_BFP"))>1)){
          g1 = ggplot(res_g1_reduced %>% subset(!is.na(padj)), aes(x=pvalue)) + 
            geom_histogram(color="darkblue", fill="lightblue", 
                           breaks = seq(0,1,by=0.01)) + 
            ggtitle(paste0(d_group, " ", cline_str, " ", g_oe, "_", ko_type))
          print(g1)
        }

        
        tag1 = sprintf("%s_%s_", g_oe, ko_type)
        colnames(res_g1) = paste0(tag1, colnames(res_g1))
        colnames(res_g1_reduced) = paste0(tag1, colnames(res_g1_reduced))      
        
        if(is.null(res_cellline_df)){
          res_cellline_df = res_g1
        }else if(!is.null(res_cellline_df)){
          stopifnot(all(rownames(res_cellline_df) == rownames(res_g1)))
          res_cellline_df = cbind(res_cellline_df, res_g1)
        }

        if(is.null(res_cellline_reduced_df)){
          res_cellline_reduced_df = res_g1_reduced
        }else if(!is.null(res_cellline_reduced_df)){
          stopifnot(all(rownames(res_cellline_reduced_df) == rownames(res_g1_reduced)))
          res_cellline_reduced_df = cbind(res_cellline_reduced_df, res_g1_reduced)
        }       

      }

    
      dim(res_cellline_df)
      res_cellline_df[1:2,]
      
      dim(res_cellline_reduced_df)
      res_cellline_reduced_df[1:2,]
      
      res_df = data.frame(gene_ID=rownames(res_cellline_df), 
                          res_cellline_df)
      dim(res_df)
      res_df[1:2,]
  
      res_reduced_gene_anno = gene_anno[match(rownames(res_cellline_reduced_df), 
                                              gene_anno$ensembl_gene_id), ]
      stopifnot(all(rownames(res_cellline_reduced_df)==res_reduced_gene_anno$ensembl_gene_id))
  
      # set gene hgnc_symbol that equal "" to NA
      hgnc_symbol_vec = res_reduced_gene_anno$hgnc_symbol
      hgnc_symbol_vec[which(hgnc_symbol_vec=="")] = NA
        
      res_reduced_df = data.frame(gene_ID=rownames(res_cellline_reduced_df), 
                                  hgnc_symbol=hgnc_symbol_vec,
                                  chr=res_reduced_gene_anno$chr, 
                                  strand=res_reduced_gene_anno$strand, 
                                  start=res_reduced_gene_anno$start, 
                                  end=res_reduced_gene_anno$end, 
                                  res_cellline_reduced_df)
      
      print("hgnc_symbol empty string and NA tables:")
      print(table(res_reduced_df$hgnc_symbol=="", useNA="ifany"))
      print(table(is.na(res_reduced_df$hgnc_symbol)))
        
      dim(res_reduced_df)
      res_reduced_df[1:2,]
      
      setting_name = unique(paste0(sample_info_dg$model_day, "_", 
#                                   sample_info_dg$condition, "_", 
                                   cline_str))
      
      fwrite(res_df, 
             sprintf("%s/%s_%s_%s_DEseq2_raw.tsv", 
                     raw_output_dir, release_name, setting_name, 
                     paste0(g_oe, "_", ko_type)), 
             sep="\t")
      fwrite(res_reduced_df, 
             sprintf("%s/%s_%s_%s_DEseq2.tsv", 
                     processed_output_dir, release_name, setting_name, 
                     paste0(g_oe, "_", ko_type)),  
             sep="\t")
    
    }
    
  }

get_index <- function(x){
  which(x$padj < 0.05 & abs(x$log2FoldChange) > log2(1.5))
}

# The four settings
settings    <- names(res_cellline)
short_names <- sub("_.*$", "", settings)

settings
short_names

# DE indices and DE gene names per setting
w_list   <- lapply(settings, function(s) get_index(res_cellline[[s]]))
names(w_list) <- settings
de_genes <- lapply(settings, function(s) rownames(res_cellline[[s]])[w_list[[s]]])
names(de_genes) <- settings

print(paste0("DE group ", d_group, " under OE gene ", g_oe, ":"))

# Per-setting DE counts
for (i in length(settings)) {
  s <- settings[i]
  short_s <- short_names[i]
  print(paste0("number of DE genes from cell line with knock out gene ", short_s, ": ",
               as.character(length(w_list[[s]]))))
}

# All pairwise overlaps (6 pairs for 4 settings)
pairs <- combn(settings, 2, simplify = FALSE)
for (p in pairs) {
  ov <- length(intersect(de_genes[[p[1]]], de_genes[[p[2]]]))
  print(paste0("number of common DE genes by cell lines with knock out gene ",
               sub("_.*$", "", p[1]), " and ", sub("_.*$", "", p[2]), ": ", as.character(ov)))
}

# UpSet plot across all 4 sets (with shortened labels)
if (max(sapply(w_list, length)) > 0) {
  names(de_genes) <- short_names
  m <- make_comb_mat(de_genes)
  g_up <- UpSet(m)
  print(g_up)
  names(de_genes) <- settings   # restore for any downstream use
}

# padj matrix for correlations
df1 <- as.data.frame(sapply(settings, function(s) res_cellline[[s]]$padj))
colnames(df1) <- paste0("padj_", settings)

# Every pairwise scatter (6 plots)
for (p in pairs) {
  x_col <- paste0("padj_", p[1])
  y_col <- paste0("padj_", p[2])
  cr <- cor(df1[[x_col]], df1[[y_col]], method = "spearman", use = "pair")

  df_plot <- data.frame(x = -log10(df1[[x_col]]),
                        y = -log10(df1[[y_col]]))
  df_plot <- df_plot[is.finite(df_plot$x) & is.finite(df_plot$y), ]

  if (nrow(df_plot) < 2) next   # nothing to plot for this pair

  g <- ggplot(df_plot, aes(x = x, y = y)) +
    geom_pointdensity() +
    scale_color_viridis() +
    xlab(paste0("-log10(padj) ", sub("_.*$", "", p[1]))) +
    ylab(paste0("-log10(padj) ", sub("_.*$", "", p[2]))) +
    ggtitle(sprintf("%s, %s, %s\n%s vs %s\nSpearman r = %.2f",
                    d_group, g_oe, ko_type, sub("_.*$", "", p[1]), sub("_.*$", "", p[2]), cr))
  print(g)
}

}
```

Save the size dataframe out
```{r fig.width = 4.5, fig.height=3.2, message=FALSE, warning=FALSE}
colnames(df_size) = c("DE_group", "cell_line", "knockout_gene_strategy", "n_ko", "n_WT")
df_size = as.data.frame(df_size)

multi_rows = which(table(df_size$knockout_gene_strategy)>1)
multi_rows

df_size[which(df_size$knockout_gene_strategy%in%names(multi_rows)),]

fwrite(df_size, 
       sprintf("%s/%s_DE_n_samples.tsv", output_dir, release_name), 
       sep="\t")

```




## DE analysis under control (BFP OE) only, between KOs and WTs

There is only one DE group for this dataset and DESeq2 model was already fit. 

Directly use the fitted model to extract the DE results for this setting. 

```{r fig.width = 4.8, fig.height=3.2, message=FALSE, warning=FALSE}

df_size = NULL

genos = setdiff(cell_lines, "WT")
genos = sort(genos)
genos

res_geno = list()

for(geno in genos){
  
  res_geno_df = NULL
  res_geno_reduced_df = NULL
    
  res_g1 = results(dds, contrast = c("ko_group", 
                                     paste0(geno, "_BFP"), paste0("WT_BFP")))
  res_g1 = signif(data.frame(res_g1), 3)
  
  # add a column to record the number of zeros per test
  test_index = which(sample_info_dg$ko_group%in%c(paste0(geno, "_BFP"), paste0("WT_BFP")))
  
  cts_dat_m_dg_test = cts_dat_m_dg[, test_index]
  n_zero = rowSums(cts_dat_m_dg_test==0)
  res_g1$n_zeros = n_zero
  
  # record the number of samples involved in the comparison
  test_ko_group_vec = sample_info_dg$ko_group[test_index]
  
  df_size = rbind(df_size, 
                  c(d_group, 
                    paste0("control"), 
                    paste0(geno, "_KO"), 
                    sum(test_ko_group_vec!="WT_BFP"), 
                    sum(test_ko_group_vec=="WT_BFP")))
  
  # prepare a processed version of output
  res_g1_reduced = res_g1[, c("log2FoldChange", "pvalue", "padj", "n_zeros")]
  res_g1_reduced$padj[which(res_g1_reduced$n_zeros>=4)] = NA
  
  if ((sum(test_ko_group_vec!=paste0(geno, "_BFP"))==1)|(sum(test_ko_group_vec==paste0("WT_BFP"))==1)){
    res_g1_reduced$padj = NA
  }
  
  res_g1_reduced = res_g1_reduced[, c("log2FoldChange", "pvalue", "padj")]
  
  res_geno[[paste0(geno, "_KO")]] = res_g1_reduced
  
  if ((sum(test_ko_group_vec!=paste0(geno, "_BFP"))>1) & (sum(test_ko_group_vec==paste0("WT_BFP"))>1)){
    g1 = ggplot(res_g1_reduced %>% subset(!is.na(padj)), aes(x=pvalue)) + 
      geom_histogram(color="darkblue", fill="lightblue", 
                     breaks = seq(0,1,by=0.01)) + 
      ggtitle(paste0(d_group, " ", "control ", geno, "_KO "))
    print(g1)
  }

  
  tag1 = sprintf("%s_%s_", geno, "KO")
  colnames(res_g1) = paste0(tag1, colnames(res_g1))
  colnames(res_g1_reduced) = paste0(tag1, colnames(res_g1_reduced))      
  
  if(is.null(res_geno_df)){
    res_geno_df = res_g1
  }else if(!is.null(res_geno_df)){
    stopifnot(all(rownames(res_geno_df) == rownames(res_g1)))
    res_geno_df = cbind(res_geno_df, res_g1)
  }

  if(is.null(res_geno_reduced_df)){
    res_geno_reduced_df = res_g1_reduced
  }else if(!is.null(res_geno_reduced_df)){
    stopifnot(all(rownames(res_geno_reduced_df) == rownames(res_g1_reduced)))
    res_geno_reduced_df = cbind(res_geno_reduced_df, res_g1_reduced)
  }       



  dim(res_geno_df)
  res_geno_df[1:2,]
  
  dim(res_geno_reduced_df)
  res_geno_reduced_df[1:2,]
  
  res_df = data.frame(gene_ID=rownames(res_geno_df), 
                      res_geno_df)
  dim(res_df)
  res_df[1:2,]

  res_reduced_gene_anno = gene_anno[match(rownames(res_geno_reduced_df), 
                                          gene_anno$ensembl_gene_id), ]
  stopifnot(all(rownames(res_geno_reduced_df)==res_reduced_gene_anno$ensembl_gene_id))

  # set gene hgnc_symbol that equal "" to NA
  hgnc_symbol_vec = res_reduced_gene_anno$hgnc_symbol
  hgnc_symbol_vec[which(hgnc_symbol_vec=="")] = NA
    
  res_reduced_df = data.frame(gene_ID=rownames(res_geno_reduced_df), 
                              hgnc_symbol=hgnc_symbol_vec,
                              chr=res_reduced_gene_anno$chr, 
                              strand=res_reduced_gene_anno$strand, 
                              start=res_reduced_gene_anno$start, 
                              end=res_reduced_gene_anno$end, 
                              res_geno_reduced_df)
  
  print("hgnc_symbol empty string and NA tables:")
  print(table(res_reduced_df$hgnc_symbol=="", useNA="ifany"))
  print(table(is.na(res_reduced_df$hgnc_symbol)))
    
  dim(res_reduced_df)
  res_reduced_df[1:2,]

  setting_name = unique(paste0(sample_info_dg$model_day, "_", 
#                                   sample_info_dg$condition, "_", 
                               "control"))
  
  fwrite(res_df, 
         sprintf("%s/%s_%s_%s_vs_WT_DEseq2_raw.tsv", 
                 raw_output_dir, release_name, setting_name, 
                 paste0(geno, "_", "KO")), 
         sep="\t")
  fwrite(res_reduced_df, 
         sprintf("%s/%s_%s_%s_vs_WT_DEseq2.tsv", 
                 processed_output_dir, release_name, setting_name, 
                 paste0(geno, "_", "KO")),  
         sep="\t")

}
  
colnames(df_size) = c("DE_group", "OE_setting", "knockout_gene_strategy", "n_ko", "n_WT")
df_size = as.data.frame(df_size)

fwrite(df_size, 
       sprintf("%s/%s_DE_control_KO_vs_WT_n_samples.tsv", output_dir, release_name), 
       sep="\t")


```





```{r}
gc()
sessionInfo()
```
