---
title: "Check the results of differential expression analysis for 2026_01_MSK_ISL1_OE"
author: "Si Liu, Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R libraries.

```{r load_libraries, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(stringr)
library(MASS)
library(RColorBrewer)
library(DESeq2)

library(viridis)
library(ggpointdensity)
library(dplyr)
library(data.table)
library(ComplexHeatmap)
library(UpSetR)
library(readxl)
theme_set(theme_classic())

library(svglite)

path = "../../MorPhiC/data/MorPhiC_internal_releases/MorPhiC_Release_January_2026/MSK_bulkRNAseq_ISL1_OE"
# path = "./data"
dat_path = file.path(path, "processed/Tables")

meta_save_path = "./data/Jan_2026"
meta_file = file.path(meta_save_path, "MSK_bulkRNAseq_ISL1_OE_meta_data.tsv")

```


## Read in meta data
```{r fig.width = 3, fig.height=3}
meta = fread(meta_file, sep="\t", data.table = FALSE)
dim(meta)

table(table(meta$clonal.label))
# table(table(meta$library_preparation.label))
table(table(meta$label))
```


## Read in gene count data and filter genes

There are redundancy in the count data file.

    The 2nd and 3rd columns are the same and only one of them should be kept. 
    Only keep the columns that show up in the sequence table

```{r fig.width = 4, fig.height=3}
cts_lanes = fread(file.path(dat_path, "genesCounts.csv"), data.table = FALSE)
dim(cts_lanes)
cts_lanes[1:2, c(1:2, (ncol(cts_lanes)-1):ncol(cts_lanes))]

names(cts_lanes)[1:6]

length(unique(colnames(cts_lanes)[2:ncol(cts_lanes)]))

# remove the third column from the cts_lanes object since it is a duplicate of the second column
cts_lanes = cts_lanes[,-3]
names(cts_lanes)[1:6]

# only keep the columns that show up in the sequence table
# summarized as the info in the sample_id column of meta data table
names_short = gsub("_L00[1-8]$", "", names(cts_lanes))
cols2kp = which(names_short %in% meta$sample_id)
length(cols2kp)
cts_lanes = cts_lanes[, c(1, cols2kp)]
dim(cts_lanes)

names_short_filterd = gsub("_L00[1-8]$", "", names(cts_lanes))

stopifnot(setequal(names_short_filterd[-1], meta$sample_id))

# Separate the gene ID column from the count columns
gene_column <- cts_lanes$Name
counts   <- as.matrix(cts_lanes[, -1])   # drop the "Name" column

# Derive sample ID from each column name by stripping "_L00X" at the end
sample_id <- sub("_L00[1-8]$", "", colnames(counts))

# Sum lane columns belonging to the same sample.
collapsed <- t(rowsum(t(counts), group = sample_id))

# shorten column names further to just "ISL1-BFP-1" etc.
colnames(collapsed) <- sub("_IGO.*", "", colnames(collapsed))

# Put the gene IDs back as a column
cts <- data.frame(Name = gene_column, collapsed, check.names = FALSE)

head(cts)
dim(cts_lanes)        # original
dim(cts)   # should have ~1/4 the columns + 1 for Name

# reorder the meta data table to match the samples in the processed count data
meta = meta[match(names(cts)[-1], meta$clonal.label),]
table(names(cts)[-1] == meta$clonal.label, useNA="ifany")

cts_dat = data.matrix(cts[,-1])
rownames(cts_dat) = cts$Name

table(meta$timepoint_value, useNA="ifany")
meta$timepoint = as.character(meta$timepoint_value)

stopifnot(length(unique(meta$timepoint)) == 1)
unique_tp = sprintf("day %s", unique(meta$timepoint))

stopifnot(length(unique(meta$model_system)) == 1)
unique_ms = unique(meta$model_system)
```

### Discard genes whose expression is 0 in more than 75% of samples

```{r fig.width = 4, fig.height=3}
model_s = meta$model_system
table(model_s, useNA="ifany")

get_q75 <- function(x){quantile(x, probs = 0.75)}

gene_info = data.frame(Name = cts$Name,
                       apply(cts_dat, 1, min), 
                       apply(cts_dat, 1, median), 
                       apply(cts_dat, 1, get_q75), 
                       apply(cts_dat, 1, max))

dim(gene_info)
gene_info[1:2,]
table(row.names(gene_info)==gene_info$Name, useNA="ifany")

names(gene_info)[2:5] = paste0(rep(c(model_s), times=4), 
                               rep(c("min", "median", "q75", "max"), each=1))
dim(gene_info)
gene_info[1:2,]

summary(gene_info[,-1])

w2kp = which(gene_info$Pancreatic_Isletq75 > 0)
cts_dat = cts_dat[w2kp,]
dim(cts_dat)

gene_info = gene_info[w2kp,]
dim(gene_info)
```

## Read in gene annotation
```{r fig.width = 4, fig.height=3}
gene_anno = fread("data/gencode_v44_primary_assembly_info.tsv")
dim(gene_anno)
gene_anno[1:2,]

table(gene_info$Name %in% gene_anno$ensembl_gene_id)
# all ensembl gene IDs are contained in the annotation
setdiff(gene_info$Name, gene_anno$ensembl_gene_id)

gene_info = merge(gene_anno, gene_info, by.x="ensembl_gene_id", by.y = "Name", 
                  all.x = FALSE, all.y = TRUE)
dim(gene_info)
gene_info[1:2,]

gene_info$gene_symbol = gene_info$hgnc_symbol

table(gene_info$gene_symbol == "")
table(is.na(gene_info$gene_symbol))

wEns = which(gene_info$gene_symbol == "" | is.na(gene_info$gene_symbol))
gene_info$gene_symbol[wEns] = gene_info$ensembl_gene_id[wEns]

table(gene_info$gene_symbol == "")
table(is.na(gene_info$gene_symbol))
```

## Prepare for reading in DE results

Use each combination of (model system, condition) as DE group. 

Run analysis for each DE group separately.

The specific DE groups can be found in the file with naming in the format:

    ${dataset_name}_DE_n_samples.tsv
    
There are also DE results under control between samples under knockout and WT settings.

```{r fig.width = 4, fig.height=3}
release_name = "2026_01_MSK_bulkRNAseq_ISL1_OE"
result_dir = file.path("results", release_name)
processed_output_dir = file.path(result_dir, "processed")

all_result_files = list.files(path=processed_output_dir, pattern=".tsv", 
                              all.files=TRUE, full.names=FALSE)
all_result_files = sort(all_result_files)

regular_DE_files = all_result_files[!grepl("KO_vs_WT", all_result_files)]
condition_DE_files = all_result_files[grepl("KO_vs_WT", all_result_files)]

extract_cores <- function(x){
  x_core = sub(paste0(release_name, "_"), "", x, fixed = TRUE)
  x_core = sub("_DEseq2.tsv", "", x_core, fixed = TRUE)
  x_split = str_split(x_core, "_")[[1]]
  x_end = length(x_split)
  x_d_group = paste(x_split[1:(x_end-3)], collapse="_")
  x_cline = paste(x_split[(x_end-2):(x_end-2)], collapse="_")
  if (x_end > 7){
    x_cline = paste(x_split[(x_end-3):(x_end-2)], collapse="_")
    x_d_group = paste(x_split[1:(x_end-4)], collapse="_")
  }
  x_gene = x_split[x_end-1]
  return(c(x_d_group, x_cline, x_gene))
}

df_file_info = NULL

for (x in regular_DE_files){
  df_file_info = rbind(df_file_info, extract_cores(x))
}

df_file_info = as.data.frame(df_file_info)
colnames(df_file_info) = c("DE_general_group", "cell_line", "gene")
df_file_info


extract_cores_cond <- function(x){
  x_core = sub(paste0(release_name, "_"), "", x, fixed = TRUE)
  x_core = sub("_DEseq2.tsv", "", x_core, fixed = TRUE)
  x_split = str_split(x_core, "_")[[1]]
  x_end = length(x_split)
  x_d_group = paste(x_split[1:(x_end-5)], collapse="_")
  x_OE = x_split[x_end-4]
  x_gene = x_split[x_end-3]
  return(c(x_d_group, x_OE, x_gene))
}

df_file_info_cond = NULL

for (x in condition_DE_files){
  df_file_info_cond = rbind(df_file_info_cond, extract_cores_cond(x))
}

df_file_info_cond = as.data.frame(df_file_info_cond)
colnames(df_file_info_cond) = c("DE_general_group", "OE_setting", "gene")
df_file_info_cond

```


## Prepare output directories for saving figure files out

```{r fig.width = 4.5, fig.height=3, message=FALSE, warning=FALSE}

df_size = NULL

figure_dir = file.path("figures", release_name)
volcano_figure_dir = file.path(figure_dir, "volcano_plots")
volcano_png_width_px = 1920
volcano_png_height_px = 1230
volcano_max_overlaps = 20
volcano_label_top_n = 20

if (!file.exists(volcano_figure_dir)){
  dir.create(volcano_figure_dir, recursive = TRUE)
}

```



## Number of DE genes and properties of ISL1

```{r fig.width = 7, fig.height=4}

fc0 = log2(1.5)
p0  = 0.05

make_result_summary <- function(file_name, comparison_name, figure_group){
  
  res = fread(file.path(processed_output_dir, file_name), data.table = FALSE)
  
  fc = paste0(comparison_name, "_log2FoldChange")
  padj = paste0(comparison_name, "_padj")
  stopifnot(all(c("gene_ID", fc, padj) %in% names(res)))
  
  data.frame(file_name = file_name,
             figure_group = figure_group,
             comparison_name = comparison_name,
             nDE = sum(abs(res[[fc]]) > fc0 & res[[padj]] < p0, na.rm = TRUE))
}

make_volcano_data <- function(file_name, comparison_name){
  
  res = fread(file.path(processed_output_dir, file_name), data.table = FALSE)
  
  fc = paste0(comparison_name, "_log2FoldChange")
  pvalue = paste0(comparison_name, "_pvalue")
  padj = paste0(comparison_name, "_padj")
  stopifnot(all(c("gene_ID", fc, pvalue, padj) %in% names(res)))
  
  res_k = res[, c("gene_ID", "hgnc_symbol", fc, pvalue, padj)]
  names(res_k) = c("gene_ID", "hgnc_symbol", "log2FoldChange", "pvalue", "padj")
  
  n_total = nrow(res_k)
  n_padj_na = sum(is.na(res_k$padj))
  n_lfc_na = sum(is.na(res_k$log2FoldChange))
  n_padj0 = sum(!is.na(res_k$padj) & is.finite(res_k$padj) & res_k$padj == 0)
  
  keep = is.finite(res_k$log2FoldChange) & !is.na(res_k$padj) &
    is.finite(res_k$padj) & res_k$padj >= 0
  res_k = res_k[keep, , drop = FALSE]
  
  if (nrow(res_k) == 0){
    return(list(data = res_k,
                diagnostics = data.frame(n_total = n_total,
                                         n_padj_na = n_padj_na,
                                         n_lfc_na = n_lfc_na,
                                         n_padj0 = n_padj0,
                                         n_plotted = nrow(res_k))))
  }
  
  res_k$gene_symbol = res_k$hgnc_symbol
  res_k$gene_symbol[is.na(res_k$gene_symbol) | res_k$gene_symbol == ""] =
    res_k$gene_ID[is.na(res_k$gene_symbol) | res_k$gene_symbol == ""]
  
  res_k$DE = "No"
  res_k$DE[which(res_k$log2FoldChange > fc0 & res_k$padj < p0)] = "Up"
  res_k$DE[which(res_k$log2FoldChange < -fc0 & res_k$padj < p0)] = "Down"
  
  res_k$log2FoldChange_plot = pmin(pmax(res_k$log2FoldChange, -3), 3)
  res_k$padj_plot = pmax(res_k$padj, .Machine$double.xmin)
  
  res_k$delabel = NA
  w_de = which(res_k$DE != "No")
  if (length(w_de) > 0) {
    ord = order(res_k$padj[w_de], decreasing = FALSE, na.last = NA)
    w_lbl = w_de[ord][seq_len(min(length(ord), volcano_label_top_n))]
    res_k$delabel[w_lbl] = res_k$gene_symbol[w_lbl]
  }
  
  list(data = res_k,
       diagnostics = data.frame(n_total = n_total,
                                n_padj_na = n_padj_na,
                                n_lfc_na = n_lfc_na,
                                n_padj0 = n_padj0,
                                n_plotted = nrow(res_k)))
}

plot_volcano <- function(res_k, plot_title){
  
  col2use = c("blue", "grey", "red")
  names(col2use) = c("Down", "No", "Up")
  
  ggplot(res_k, aes(x = log2FoldChange_plot, y = -log10(padj_plot),
                    col = DE, label = delabel)) +
    geom_point() +
    geom_text_repel(point.padding = 0.5,
                    min.segment.length = 0,
                    size = 3,
                    box.padding = 0.5,
                    max.overlaps = volcano_max_overlaps,
                    show.legend = FALSE) +
    scale_color_manual(values = col2use) +
    labs(title = plot_title,
         x = "log2FoldChange",
         y = "-log10(padj)")
}

regular_de_info = df_file_info
regular_de_info$file_name = regular_DE_files
regular_de_info$comparison_name = paste0(regular_de_info$gene, "_OE")
regular_de_info$background = regular_de_info$cell_line

regular_de_info

df_regular = do.call(rbind, lapply(seq_len(nrow(regular_de_info)), function(k){
  make_result_summary(file_name = regular_de_info$file_name[k],
                      comparison_name = regular_de_info$comparison_name[k],
                      figure_group = regular_de_info$background[k])
}))

df_regular$background = factor(df_regular$figure_group,
                               levels = regular_de_info$background)
df_regular

p_regular_nDE = ggplot(df_regular, aes(x = background, y = nDE, fill = background)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = nDE), vjust = -0.35, size = 3) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "Number of DE genes from ISL1 OE",
       x = "Cell line / background",
       y = "Number of DE genes") +
  theme(legend.position = "none")

print(p_regular_nDE)

```


### ISL1 OE effect on ISL1

```{r fig.width = 6, fig.height=4}

isl1_gene_ids = gene_info$ensembl_gene_id[which(gene_info$gene_symbol == "ISL1")]
isl1_gene_ids
stopifnot(length(isl1_gene_ids) == 1)

df_isl1 = do.call(rbind, lapply(seq_len(nrow(regular_de_info)), function(k){
  
  res = fread(file.path(processed_output_dir, regular_de_info$file_name[k]),
              data.table = FALSE)
  
  fc = paste0(regular_de_info$comparison_name[k], "_log2FoldChange")
  padj = paste0(regular_de_info$comparison_name[k], "_padj")
  w_isl1 = which(res$gene_ID == isl1_gene_ids)
  stopifnot(length(w_isl1) == 1)
  
  data.frame(background = regular_de_info$background[k],
             log2FoldChange = res[[fc]][w_isl1],
             padj = res[[padj]][w_isl1])
}))

df_isl1$background = factor(df_isl1$background,
                            levels = regular_de_info$background)
df_isl1

p_isl1_effect = ggplot(df_isl1, aes(x = log2FoldChange,
                                    y = -log10(padj),
                                    color = background)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = background),
                  point.padding = 0.5,
                  min.segment.length = 0,
                  size = 3,
                  box.padding = 0.5,
                  max.overlaps = 20,
                  show.legend = FALSE) +
  scale_color_brewer(palette = "Dark2") +
  geom_hline(yintercept = -log10(p0), color = "grey", linetype = "dashed") +
  geom_vline(xintercept = c(-fc0, fc0), color = "grey", linetype = "dashed") +
  labs(title = "ISL1 OE effect on ISL1",
       x = "ISL1 log2 fold change",
       y = "-log10(adjusted p-value)")

print(p_isl1_effect)

```


### Control knockout versus WT

```{r fig.width = 5, fig.height=3.5}

condition_de_info = df_file_info_cond
condition_de_info$file_name = condition_DE_files
condition_de_info$comparison_name = paste0(condition_de_info$gene, "_KO")

condition_de_info

df_condition = do.call(rbind, lapply(seq_len(nrow(condition_de_info)), function(k){
  make_result_summary(file_name = condition_de_info$file_name[k],
                      comparison_name = condition_de_info$comparison_name[k],
                      figure_group = condition_de_info$gene[k])
}))

df_condition$gene = factor(df_condition$figure_group,
                           levels = condition_de_info$gene)
df_condition

p_condition_nDE = ggplot(df_condition, aes(x = gene, y = nDE, fill = gene)) +
  geom_col(width = 0.75) +
  geom_text(aes(label = nDE), vjust = -0.35, size = 3) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "Number of DE genes under control",
       subtitle = "KO versus WT",
       x = "Knockout gene",
       y = "Number of DE genes") +
  theme(legend.position = "none")

print(p_condition_nDE)

```


## Generate volcano plots

### ISL1 OE under each background

```{r fig.width = 5, fig.height=3.5}

for(k in seq_len(nrow(regular_de_info))){
  
  volcano_k = make_volcano_data(file_name = regular_de_info$file_name[k],
                                comparison_name = regular_de_info$comparison_name[k])
  res_k = volcano_k$data
  diag_k = volcano_k$diagnostics
  
  print(table(res_k$DE))
  cat(regular_de_info$comparison_name[k],
      " @ ",
      regular_de_info$background[k],
      ": genes=",
      diag_k$n_total,
      ", padj_NA=",
      diag_k$n_padj_na,
      ", log2FC_NA=",
      diag_k$n_lfc_na,
      ", padj_0=",
      diag_k$n_padj0,
      ", plotted=",
      diag_k$n_plotted,
      "\n",
      sep = "")
  if (nrow(res_k) == 0) next
  
  plot_title = paste0(unique_tp, " ", regular_de_info$background[k], "_", regular_de_info$comparison_name[k], " vs ",
                      regular_de_info$background[k], "_", "control")
  g2 = plot_volcano(res_k, plot_title)
  print(g2)
  
  #g3 = plot_volcano(res_k, "") + theme(
  g3 = plot_volcano(res_k, "") + ggtitle(NULL) + theme(
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA)
  )
  
  saved_figure_name = paste0(regular_de_info$background[k], "_",
                             regular_de_info$comparison_name[k],
                             "_volcano.png")
  ggsave(filename = file.path(volcano_figure_dir, saved_figure_name),
         plot = g3,
         device = grDevices::png,
         width = volcano_png_width_px,
         height = volcano_png_height_px,
         units = "px",
         bg = "transparent")
}

```


### Control knockout versus WT

```{r fig.width = 5, fig.height=3.5}

for(k in seq_len(nrow(condition_de_info))){
  
  volcano_k = make_volcano_data(file_name = condition_de_info$file_name[k],
                                comparison_name = condition_de_info$comparison_name[k])
  res_k = volcano_k$data
  diag_k = volcano_k$diagnostics
  
  print(table(res_k$DE))
  cat(condition_de_info$comparison_name[k],
      " @ ",
      condition_de_info$OE_setting[k],
      ": genes=",
      diag_k$n_total,
      ", padj_NA=",
      diag_k$n_padj_na,
      ", log2FC_NA=",
      diag_k$n_lfc_na,
      ", padj_0=",
      diag_k$n_padj0,
      ", plotted=",
      diag_k$n_plotted,
      "\n",
      sep = "")
  if (nrow(res_k) == 0) next
  
  plot_title = paste0(unique_tp, " ", 
                      condition_de_info$comparison_name[k], "_", 
                      condition_de_info$OE_setting[k], " vs ",
                      "WT", "_", condition_de_info$OE_setting[k])
  g2 = plot_volcano(res_k, plot_title)
  print(g2)
  
  g3 = plot_volcano(res_k, "") + ggtitle(NULL) + theme(
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA)
  )
  
  saved_figure_name = paste0(condition_de_info$OE_setting[k], "_",
                             condition_de_info$comparison_name[k],
                             "_KO_vs_WT_volcano.png")
  ggsave(filename = file.path(volcano_figure_dir, saved_figure_name),
         plot = g3,
         device = grDevices::png,
         width = volcano_png_width_px,
         height = volcano_png_height_px,
         units = "px",
         bg = "transparent")
}

```


```{r}
gc()
sessionInfo()
```
