---
title: "Differential expression analysis for 2026-01 release MSK_bulkRNAseq_ISL1_OE: control (BFP_OE) KO vs WT model"
author: "Si Liu, Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file use the control (BFP_OE) samples from MSK_bulkRNAseq_ISL1_OE
data to do DE analysis between each gene KO and WT. 

* keep only BFP OE samples, which are the controls for OE,
* fit one DESeq2 model using all BFP OE samples,
* extract only the three control KO vs WT contrasts.

## R libraries

```{r load_libraries, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(stringr)
library(MASS)
library(RColorBrewer)
library(DESeq2)

library(viridis)
library(ggpointdensity)
library(dplyr)
library(data.table)
library(ComplexHeatmap)
library(UpSetR)
library(readxl)
theme_set(theme_classic())

path = "../../MorPhiC/data/MorPhiC_internal_releases/MorPhiC_Release_January_2026/MSK_bulkRNAseq_ISL1_OE"
dat_path = file.path(path, "processed/Tables")

meta_save_path = "./data/Jan_2026"
meta_file = file.path(meta_save_path, "MSK_bulkRNAseq_ISL1_OE_meta_data.tsv")

```

## Load processed metadata

```{r load_metadata}
meta = fread(meta_file, sep="\t", data.table = FALSE)
dim(meta)
```

## Read count data and filter genes

```{r read_counts}
cts_lanes = fread(file.path(dat_path, "genesCounts.csv"), data.table = FALSE)
dim(cts_lanes)

# remove the third column from the cts_lanes object since it is a duplicate
# of the second column
cts_lanes = cts_lanes[,-3]

# only keep the columns that show up in the sequence table
names_short = gsub("_L00[1-8]$", "", names(cts_lanes))
cols2kp = which(names_short %in% meta$sample_id)
cts_lanes = cts_lanes[, c(1, cols2kp)]

names_short_filterd = gsub("_L00[1-8]$", "", names(cts_lanes))
stopifnot(setequal(names_short_filterd[-1], meta$sample_id))

gene_column <- cts_lanes$Name
counts <- as.matrix(cts_lanes[, -1])
sample_id <- sub("_L00[1-8]$", "", colnames(counts))

# Sum lane columns belonging to the same sample.
collapsed <- t(rowsum(t(counts), group = sample_id))
colnames(collapsed) <- sub("_IGO.*", "", colnames(collapsed))

cts <- data.frame(Name = gene_column, collapsed, check.names = FALSE)

# reorder the metadata table to match the samples in the processed count data
meta = meta[match(names(cts)[-1], meta$clonal.label),]
stopifnot(all(names(cts)[-1] == meta$clonal.label))

cts_dat = data.matrix(cts[,-1])
rownames(cts_dat) = cts$Name

meta$timepoint = as.character(meta$timepoint_value)

table(meta$OE_status)

BFP_kept = which(meta$OE_status == "BFP")
meta = meta[BFP_kept,]
cts_dat = cts_dat[, BFP_kept]

table(meta$OE_status, useNA="ifany")
table(colnames(cts_dat)==meta$clonal.label, useNA="ifany")

```

### Read gene annotation

```{r read_gene_annotation}
gene_anno = fread("data/gencode_v44_primary_assembly_info.tsv")
stopifnot(all(rownames(cts_dat) %in% gene_anno$ensembl_gene_id))
```

### Initial gene filter and sample QC covariates

```{r initial_gene_filter}
model_s = meta$model_system
table(model_s)
model_s = unique(model_s)

get_q75 <- function(x){quantile(x, probs = 0.75)}

gene_info = data.frame(Name = cts$Name,
                       apply(cts_dat, 1, min),
                       apply(cts_dat, 1, median),
                       apply(cts_dat, 1, get_q75),
                       apply(cts_dat, 1, max))

names(gene_info)[2:5] = paste0(rep(c(model_s), times=4),
                               rep(c("min", "median", "q75", "max"), each=1))

w2kp = which(gene_info$Pancreatic_Isletq75 > 0)
cts_dat = cts_dat[w2kp,]
gene_info = gene_info[w2kp,]

meta$read_depth = colSums(cts_dat)/1e6
meta$q75 = apply(cts_dat, 2, quantile, probs = 0.75)

```

## Prepare output directories

```{r prepare_output_directories}
release_name = "2026_01_MSK_bulkRNAseq_ISL1_OE"

output_dir =  file.path("results", release_name)
raw_output_dir = file.path(output_dir, "raw")
processed_output_dir = file.path(output_dir, "processed")

if (!file.exists(raw_output_dir)){
  dir.create(raw_output_dir, recursive = TRUE)
}

if (!file.exists(processed_output_dir)){
  dir.create(processed_output_dir, recursive = TRUE)
}
```

## DE analysis under control only, fitting one model using all BFP_OE samples

```{r control_bfp_only_model, fig.width = 4.8, fig.height=3.2, message=FALSE, warning=FALSE}
unique_model_ss = unique(meta$model_system)
unique_model_ss = sort(unique_model_ss)

# there is only one model system in this dataset
model1 = unique_model_ss
print(model1)

sample2kp = which(meta$model_system == model1)
cts_dat_m = cts_dat[, sample2kp]
meta_m = meta[sample2kp,]
stopifnot(all(meta_m$clonal.label == colnames(cts_dat_m)))

cols2kp = c("clonal.label", "model_system", "cell_line", "OE_status",
            "q75", "timepoint")
sample_info = meta_m[, cols2kp]
colnames(sample_info)[which(colnames(sample_info) == "timepoint")] = "day"

sample_info$ko_group = paste(sample_info$cell_line,
                             sample_info$OE_status, sep = "_")
sample_info$model_day = paste0(sample_info$model_system,
                               "_day_",
                               as.character(sample_info$day))
sample_info$de_grp = sample_info$model_day

sorted_de_grps = sort(unique(sample_info$de_grp))
stopifnot(length(sorted_de_grps) == 1)
d_group = sorted_de_grps

dg_index = which(sample_info$de_grp == d_group)
cts_dat_m_dg = cts_dat_m[, dg_index]
sample_info_dg = sample_info[dg_index,]
stopifnot(all(sample_info_dg$de_grp == d_group))
stopifnot(all(sample_info_dg$clonal.label == colnames(cts_dat_m_dg)))

# Keep only BFP OE samples, which are the controls for OE.
bfp_index = which(sample_info_dg$OE_status == "BFP")
cts_dat_bfp = cts_dat_m_dg[, bfp_index]
sample_info_bfp = sample_info_dg[bfp_index,]
stopifnot(all(sample_info_bfp$OE_status == "BFP"))
stopifnot(all(sample_info_bfp$clonal.label == colnames(cts_dat_bfp)))

print("-----------------------------------")
print(sprintf("BFP-only DESeq2 model: %s control KO vs WT", d_group))
print(table(sample_info_bfp$ko_group))
print("-----------------------------------")

# Do the same two expression filters as the original workflow, but within only
# BFP OE samples.
dg_q75 = apply(cts_dat_bfp, 1, quantile, probs = 0.75)
cts_dat_bfp = cts_dat_bfp[dg_q75 > 0,]

print(sprintf("After filtering by gene expression q75, the number of genes reduces from %d to %d",
              length(dg_q75), nrow(cts_dat_bfp)))

n_pos = rowSums(cts_dat_bfp > 0)
cts_dat_bfp = cts_dat_bfp[n_pos >= 4,]

if (length(n_pos) == nrow(cts_dat_bfp)){
  print("After filtering by nonzero expression in at least 4 samples, the number of genes does not change.")
}else{
  print(sprintf("After filtering by nonzero expression in at least 4 samples, the number of genes reduces from %d to %d",
                length(n_pos), nrow(cts_dat_bfp)))
}

# Update q75 using only genes retained for this BFP-only model.
sample_info_bfp$q75 = apply(cts_dat_bfp, 2, quantile, probs = 0.75)
sample_info_bfp$log_q75 = log(sample_info_bfp$q75)
sample_info_bfp$log_q75 = sample_info_bfp$log_q75 - mean(sample_info_bfp$log_q75)

sample_info_bfp$ko_group = factor(sample_info_bfp$ko_group)
sample_info_bfp$ko_group = relevel(sample_info_bfp$ko_group, ref = "WT_BFP")
rownames(sample_info_bfp) = colnames(cts_dat_bfp)

dds = DESeqDataSetFromMatrix(cts_dat_bfp,
                             colData = sample_info_bfp,
                             design = ~ ko_group + log_q75)

start.time <- Sys.time()
dds = DESeq(dds)
end.time <- Sys.time()
print(end.time - start.time)

res_rd = as.data.frame(results(dds, name = "log_q75"))
dim(res_rd)
res_rd[1:2,]

table(is.na(res_rd$padj))

g0 = ggplot(res_rd %>% subset(!is.na(padj)), aes(x=pvalue)) + 
  geom_histogram(color="darkblue", fill="lightblue", 
                 breaks = seq(0,1,by=0.01)) + 
  ggtitle(paste0(d_group, " read depth"))
print(g0)
  
## Rerun the analysis without read-depth if it is not significant for 
## most genes, and the number of samples is small
## i.e., proportion of non-null in the distribution is smaller than 0.01
## (this needs to be restricted to the genes with not NA adjusted pvalue)
## or if smaller than 0.1 and the number of samples is not greater than 6

pi_1_rd = max(0, mean(res_rd$pvalue[!is.na(res_rd$padj)] < 0.05) - 0.05)
pi_1_rd = max(pi_1_rd, 0, 1 - 2*mean(res_rd$pvalue[!is.na(res_rd$padj)] > 0.5))

print(sprintf("prop of non-null for rd: %.2e", pi_1_rd))

if(pi_1_rd < 0.01 || (ncol(cts_dat_bfp) <= 6 && pi_1_rd < 0.1 )){
  
  print("rerun DESeq2 without read depth")
  
  include_rd = 0

  dds = DESeqDataSetFromMatrix(cts_dat_bfp, colData=sample_info_bfp, 
                                ~ ko_group)

  start.time <- Sys.time()
  dds = DESeq(dds)
  end.time <- Sys.time()
  
  time.taken <- end.time - start.time
  print(time.taken)
  
}else{
  include_rd = 1
}


df_size = NULL
res_geno = list()
cell_lines = unique(sample_info_bfp$cell_line)
genos = setdiff(cell_lines, "WT")
genos = sort(genos)

for(geno in genos){

  res_g1 = results(dds, contrast = c("ko_group",
                                     paste0(geno, "_BFP"),
                                     "WT_BFP"))
  res_g1 = signif(data.frame(res_g1), 3)

  test_index = which(sample_info_bfp$ko_group %in% c(paste0(geno, "_BFP"),
                                                     "WT_BFP"))
  cts_dat_test = cts_dat_bfp[, test_index]
  n_zero = rowSums(cts_dat_test == 0)
  res_g1$n_zeros = n_zero

  test_ko_group_vec = sample_info_bfp$ko_group[test_index]
  df_size = rbind(df_size,
                  c(d_group,
                    "control",
                    paste0(geno, "_KO"),
                    sum(test_ko_group_vec != "WT_BFP"),
                    sum(test_ko_group_vec == "WT_BFP")))

  res_g1_reduced = res_g1[, c("log2FoldChange", "pvalue", "padj", "n_zeros")]
  res_g1_reduced$padj[which(res_g1_reduced$n_zeros >= 4)] = NA

  if ((sum(test_ko_group_vec != "WT_BFP") == 1) |
      (sum(test_ko_group_vec == "WT_BFP") == 1)){
    res_g1_reduced$padj = NA
  }

  res_g1_reduced = res_g1_reduced[, c("log2FoldChange", "pvalue", "padj")]
  res_geno[[paste0(geno, "_KO")]] = res_g1_reduced

  if ((sum(test_ko_group_vec != "WT_BFP") > 1) &
      (sum(test_ko_group_vec == "WT_BFP") > 1)){
    g1 = ggplot(res_g1_reduced %>% subset(!is.na(padj)), aes(x = pvalue)) +
      geom_histogram(color = "darkblue", fill = "lightblue",
                     breaks = seq(0, 1, by = 0.01)) +
      ggtitle(paste0(d_group, " control ", geno, "_KO"))
    print(g1)
  }

  tag1 = sprintf("%s_%s_", geno, "KO")
  colnames(res_g1) = paste0(tag1, colnames(res_g1))
  colnames(res_g1_reduced) = paste0(tag1, colnames(res_g1_reduced))

  res_df = data.frame(gene_ID = rownames(res_g1),
                      res_g1)

  res_reduced_gene_anno = gene_anno[match(rownames(res_g1_reduced),
                                          gene_anno$ensembl_gene_id),]
  stopifnot(all(rownames(res_g1_reduced) == res_reduced_gene_anno$ensembl_gene_id))

  # set gene hgnc_symbol that equal "" to NA
  hgnc_symbol_vec = res_reduced_gene_anno$hgnc_symbol
  hgnc_symbol_vec[which(hgnc_symbol_vec == "")] = NA

  res_reduced_df = data.frame(gene_ID = rownames(res_g1_reduced),
                              hgnc_symbol = hgnc_symbol_vec,
                              chr = res_reduced_gene_anno$chr,
                              strand = res_reduced_gene_anno$strand,
                              start = res_reduced_gene_anno$start,
                              end = res_reduced_gene_anno$end,
                              res_g1_reduced)
  
  print("hgnc_symbol empty string and NA tables:")
  print(table(res_reduced_df$hgnc_symbol=="", useNA="ifany"))
  print(table(is.na(res_reduced_df$hgnc_symbol)))
  
  setting_name = unique(paste0(sample_info_bfp$model_day, "_control"))

  fwrite(res_df,
         sprintf("%s/%s_%s_%s_vs_WT_DEseq2_raw.tsv",
                 raw_output_dir, release_name, setting_name,
                 paste0(geno, "_KO")),
         sep = "\t")

  fwrite(res_reduced_df,
         sprintf("%s/%s_%s_%s_vs_WT_DEseq2.tsv",
                 processed_output_dir, release_name, setting_name,
                 paste0(geno, "_KO")),
         sep = "\t")
}

colnames(df_size) = c("DE_group", "OE_setting", "knockout_gene_strategy", "n_ko", "n_WT")
df_size = as.data.frame(df_size)

fwrite(df_size,
       sprintf("%s/%s_DE_control_KO_vs_WT_n_samples.tsv", output_dir, release_name),
       sep = "\t")
```

```{r}
gc()
sessionInfo()
```
